Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-08-14 analysis/paper/paper.qmd

```{r}
#| echo: false
#| label: fig-umap-by-category
#| include: false
#| 
category_colors <- c(
  "ideal model" = "black",
  "spheroid" = "#e66101",
  "subspheroid" = "#fdb863",
  "multifacial" = "#b2abd2",
  "polyhedron" = "#5e3c99"
)

ggplot(umap_data %>% 
         mutate(category =
                if_else(category == "idealmodel",
                       "ideal model",
                       category)) %>% 
         mutate(category_lower =
                if_else(category_lower == "idealmodel",
                       "ideal model",
                       category_lower)), 
       aes(x = x, y = y)) +
  geom_point(aes(color = category_lower, 
                 size = point_size), 
             alpha = 0.9, 
             shape = 16) +
  scale_color_manual(values = category_colors, 
                     name = "Category") +
  scale_size_identity() +
  coord_cartesian(xlim = x_limits_expanded, 
                  ylim = y_limits_expanded)+ #确保范围一致
  theme_minimal(base_size = 14) +
  theme(legend.position = c(.2,.75),
         legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
         legend.box.background = element_rect(color = "black"),
         legend.box.margin = margin(0, 0, 0, 0)) +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  ) 
```

@@

cluster_colors <- RColorBrewer::brewer.pal(4, "Set2")
library(ggforce)

ggplot(umap_data %>%  
         mutate(category_lower =
                if_else(category_lower == "idealmodel",
                       "ideal model",
                       category_lower)), 
       aes(x = x, y = y)) +
  geom_point(
    data = filter(umap_data, is_idealmodel == "no"),
    aes(color = cluster),
    alpha = 0.8,
    size = 2.5
  ) +
  geom_point(
    data = filter(umap_data, 
                  is_idealmodel == "yes"),
    aes(color = cluster),
    size = 3.5,
    show.legend = FALSE
  ) +
  geom_mark_ellipse(aes(color = cluster, 
                        label = category)) +
  #stat_ellipse(
  #  aes(color = cluster),
  #  type = "norm",
  #  level = 0.95,
  #  linetype = "dashed",
  #  show.legend = FALSE
  #) +
  scale_color_manual(
    values = cluster_colors,
    name = "Cluster"
  ) +
  coord_cartesian(xlim = x_limits_expanded, 
                  ylim = y_limits_expanded) +  # 保证坐标一致
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  )
```

@@

```{r, include=FALSE}
#| label: fig-UMAP-SPHARM-clusters
#| fig-cap: "UMAP plot showing distinct clusters of shapes derived from SPHARM"

library(here)
knitr::include_graphics(here("analysis/paper/figures/UMAP_SPHARM_clusters.png"))

```{r, include=FALSE}
#| label: fig-UMAP-SPHARM-samples
#| fig-cap: "UMAP plot of SPHARM-derived shape features colored by types"

library(here)
knitr::include_graphics(here("analysis/paper/figures/UMAP_SPHARM_samples.png"))

```{r}

#| label: fig-UMAP-combined
#| fig-cap: "A. UMAP plot showing distinct clusters of shapes derived from SPHARM. B. UMAP plot of SPHARM-derived shape features colored by types"
#| echo: false

@@

p_panel_spharm <- plot_grid(
  ggdraw() + draw_image(here("analysis/paper/figures/UMAP_SPHARM_clusters.png")),
  ggdraw() + draw_image(here("analysis/paper/figures/UMAP_SPHARM_samples.png")),
  labels = "AUTO",
  ncol = 1
)

>>>>

```{r}
#| echo: false
#| label: fig-umap-by-category
#| include: false

category_colors <- c(
  "ideal model" = "black",
  "spheroid" = "#e66101",
  "subspheroid" = "#fdb863",
  "multifacial" = "#b2abd2",
  "polyhedron" = "#5e3c99"
)

p_umap_by_category <- 
ggplot(umap_data %>% 
         mutate(category =
                if_else(category == "idealmodel",
                       "ideal model",
                       category)) %>% 
         mutate(category_lower =
                if_else(category_lower == "idealmodel",
                       "ideal model",
                       category_lower)), 
       aes(x = x, y = y)) +
  geom_point(aes(color = category_lower, 
                 size = point_size), 
             alpha = 0.9, 
             shape = 16) +
  scale_color_manual(values = category_colors, 
                     name = "Category") +
  scale_size_identity() +
  coord_cartesian(xlim = x_limits_expanded, 
                  ylim = y_limits_expanded)+ #确保范围一致
  theme_minimal(base_size = 14) +
  theme(legend.position = c(.2,.75),
         legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
         legend.box.background = element_rect(color = "black"),
         legend.box.margin = margin(0, 0, 0, 0)) +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  )

# TODO add the ideal shape graphics to p_umap_by_cluster

ggsave(plot = p_umap_by_category,
       filename = here("analysis/paper/figures/umap_by_category.png"))

@@

cluster_colors <- RColorBrewer::brewer.pal(4, "Set2")

library(ggforce)
library(ggrepel)

umap_data_to_plot <- 
umap_data %>%  
         mutate(category_lower =
                if_else(category_lower == "idealmodel",
                       "ideal model",
                       category_lower),
                filename = case_when(
                  filename == "Rounded_cube" ~ "Cube",
                  filename == "Discoid_like" ~ "Discoid-like",
                 .default = filename)
                )

p_umap_by_cluster <- 
ggplot(umap_data_to_plot, 
       aes(x = x, y = y)) +
  geom_point(
    data = filter(umap_data_to_plot, 
                  is_idealmodel == "no"),
    aes(color = cluster),
    alpha = 0.8,
    size = 2.5
  ) +
  geom_point(
    data = filter(umap_data_to_plot, 
                  is_idealmodel == "yes"),
    aes(color = cluster),
    size = 5,
    shape = 3,
    show.legend = FALSE
  ) +
  geom_mark_ellipse(aes(color = cluster)) +
  geom_text_repel(data = filter(umap_data_to_plot, 
                  is_idealmodel == "yes",
                  filename == "Sphere"),
    aes(color = cluster,
        label = filename),
    size = 5,
    hjust = 0,
    xlim = c(3, NA),
    ylim = c(2.5, NA),
    nudge_x = 3,
    box.padding = 0.5,
    nudge_y = 1,
    segment.curvature = -0.1,
    segment.ncp = 5,
    segment.angle = 20) +
    geom_text_repel(data = filter(umap_data_to_plot, 
                  is_idealmodel == "yes",
                  filename == "Cube"),
    aes(color = cluster,
        label = filename),
    size = 5,
    hjust = 0,
    xlim = c(3, NA),
    ylim = c(1, NA),
    force = 5,
    nudge_x = 3,
    box.padding = 0.5,
    nudge_y = -5) +
    geom_text_repel(data = filter(umap_data_to_plot, 
                  is_idealmodel == "yes",
                  filename %in% c("Discoid-like", 
                                  "Box",
                                  "Ellipsoid")),
    aes(color = cluster,
        label = filename),
    size = 5,
    nudge_x = 0,
    box.padding = 0.5,
    nudge_y = -2,
    ylim = c(4, 4)) +
  scale_color_manual(
    values = cluster_colors,
    name = "Cluster"
  ) +
  coord_cartesian(xlim = x_limits_expanded, 
                  ylim = y_limits_expanded) +  # 保证坐标一致
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  )

# TODO add the ideal shape graphics to p_umap_by_cluster

ggsave(plot = p_umap_by_cluster,
       filename = here("analysis/paper/figures/umap_by_cluster.png"))
```

@@

```{r}
#| label: fig-UMAP-combined
#| fig-cap: "A. UMAP plot showing distinct clusters of shapes derived from SPHARM. B. UMAP plot of SPHARM-derived shape features colored by types"
#| echo: false

@@

p_panel_spharm <- plot_grid(
  ggdraw() + draw_image(here("analysis/paper/figures/umap_by_cluster.png")),
  ggdraw() + draw_image(here("analysis/paper/figures/umap_by_category.png")),
  labels = "AUTO",
  ncol = 1
)
